---
title: "Capostone Project"
author: "Erik Emilsson"
date: "20 December 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*This project is part of the Google Data Analytics Certificate on Coursera and the dataset and the fictional problem comes from Track 1 How does a bike-share navigate speedy business.*

# Business task
Cyclist is a bike/share program with over 5,800 bikes and 600 docking stations. Currently there are two types of riders: 1) **casual riders** (those who buy single-ride or full-day passes), and 2) **annual members a.k.a. Cyclistic members**. The goal is to *maximize the number of annual memberships for the company by converting more casual riders to annual members*. Lily Moreno (director of marketing) and Cyclistics finance analyst have concluded that that annual members are more profitable than casual riders. Before this, Cyclistic have focused on building general awareness and appealing to broad consumer segments.

Currently 8 percent of riders use assistive options (e.g. reclining bikes, hand tricycles, and cargo bikes) while the rest use traditional bikes. Those who use traditional bikes are more likely to be casual riders, but about 30 percent of them use the service to commute to work each day.

Of the the following three questions that will guide the larger scope of the future marketing program, this report will only answer the first question:

- <mark>**How do annual members and casual riders use Cyclistic bikes differently?** </mark>
- Why would casual riders buy Cyclistic annual memberships?
- How can Cyclistic use digital media to influence casual riders to become members?

The stakeholders are 
- Lily Moreno, director of marketing and my manager,
- The Cyclistic marketing analytics team, a team of data analysts of which I am a part of. The team supports the Cyclistic marketing strategy with data. 
- The Cyclistic executive team the will decide whether to approve the recommended marketing program

I am using R in R-markdown for data import, cleaning and visualization as well as the final reporting. R-markdown can do all these steps and having everyting in oine place keeps track of the assumptions and steps of the analysis for transparency and reporoducibilty.

# Exploratory data analysis

The dataset that will be used to answer the business task is Cyclstic's historical trip data for the previous 12 months. The dataset was downloaded in a zip file with a folder for each month, each containing data in CSV-format. The data ranges from 202110 to 202209, using ISO date notation (YYYYMMDD).
## Load packages

```{r}
#import package(s)
library(tidyverse)
library(skimr)
```

## Import data and append data
```{r}
#read all the CSV files of trip data for the past 12 months as dataframes in R
df_202110 <- read_csv("202110-divvy-tripdata.csv")
df_202111 <- read_csv("202111-divvy-tripdata.csv")
df_202112 <- read_csv("202112-divvy-tripdata.csv")
df_202201 <- read_csv("202201-divvy-tripdata.csv")
df_202202 <- read_csv("202202-divvy-tripdata.csv")
df_202203 <- read_csv("202203-divvy-tripdata.csv")
df_202204 <- read_csv("202204-divvy-tripdata.csv")
df_202205 <- read_csv("202205-divvy-tripdata.csv")
df_202206 <- read_csv("202206-divvy-tripdata.csv")
df_202207 <- read_csv("202207-divvy-tripdata.csv")
df_202208 <- read_csv("202208-divvy-tripdata.csv")
df_202209 <- read_csv("202209-divvy-publictripdata.csv")
```

## Inspect & clean data

Using the head function I checked that the headers of each of the dataframes is the same, meaning that appending all of them is possible and will make the exploratory data analysis a lot easier since it will only be done once instead of twelve times.

```{r}
#appends all the dataframes above into one single dataframe
df <- rbind(df_202110, df_202111,df_202112, df_202201, df_202202, df_202203, df_202204, df_202205, df_202206, df_202207, df_202208, df_202209)
```

The headers are made of 13 columns, which contain information such as: 
* unique rider IDs, 
* type of bike, 
* start & end times/station name/ID/coordinates, and
* type of member/rider.

Since the goal is to understand how members and caual riders ride Cyclistic bikes differently, it is reasonable to expect that the last column will be used to find correlations with other data in the dataframe.

There are some missing values in the first few rows. Using other functions should give more insight into how often they occur and for what columns.

```{r}
head(df)
```

The glimpse function shows that there are over five million rows.

```{r}
glimpse(df)
```
Using the skimr package function skim_without_charts more information about the occurances of values in the different rows can be optained (n_unique). There is also information on mising values for some of the columns. The ones 

```{r}
skim_without_charts(df_202110)
```


```{r}
skim_without_charts(df)
```



