---
title: "Capostone Project"
author: "Erik Emilsson"
date: "20 December 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*This project is part of the Google Data Analytics Certificate on Coursera and the dataset and the fictional problem comes from Track 1 How does a bike-share navigate speedy business. I am using R in R-markdown for data import, cleaning and visualization as well as the final reporting. R-markdown can do all these steps and having everything in one place keeps track of the assumptions and steps of the analysis for transparency and reporoducibilty.*

The project will be split up into three parts:
1.Business Task
2.

# 1.Business task
Cyclist is a bike/share program with over 5,800 bikes and 600 docking stations. Currently there are two types of riders: 

1) **casual riders** (those who buy single-ride or full-day passes), and 
2) **annual members a.k.a. Cyclistic members**. 

The goal is to *maximize the number of annual memberships for the company by converting more casual riders to annual members*. Lily Moreno (director of marketing) and the Cyclistic finance analysts have concluded that that annual members are more profitable than casual riders. Before this, Cyclistic have focused on building general awareness and appealing to broad consumer segments.

Currently 8 percent of riders use assistive options (e.g. reclining bikes, hand tricycles, and cargo bikes) while the rest use traditional bikes. Those who use traditional bikes are more likely to be casual riders, but about 30 percent of them use the service to commute to work each day.

Of the the following three questions that will guide the larger scope of the future marketing program, this report will only answer the first question:

* <mark>**How do annual members and casual riders use Cyclistic bikes differently?** </mark>
* Why would casual riders buy Cyclistic annual memberships?
* How can Cyclistic use digital media to influence casual riders to become annual members?

The stakeholders are:

* Lily Moreno, director of marketing and my manager,
* The Cyclistic marketing analytics team, a team of data analysts of which I am a part of. The team supports the Cyclistic marketing strategy with data. 
* The Cyclistic executive team the will decide if the recommended marketing program is approved.

# 2.Exploratory data analysis

The dataset that will be used to answer the business task is Cyclstic's historical trip data for the previous 12 months. The dataset was downloaded in a zip file with a folder for each month, each containing data in CSV-format. The data ranges from 202110 to 202209, using ISO date notation (YYYYMMDD).

## Load packages

```{r}
#import package(s)
library(tidyverse)
library(skimr)
library(janitor)
```

## Import data and append data
```{r, message=FALSE, echo=FALSE}
#read all the CSV files of trip data for the past 12 months as dataframes in R
df_202110 <- read_csv("202110-divvy-tripdata.csv")
df_202111 <- read_csv("202111-divvy-tripdata.csv")
df_202112 <- read_csv("202112-divvy-tripdata.csv")
df_202201 <- read_csv("202201-divvy-tripdata.csv")
df_202202 <- read_csv("202202-divvy-tripdata.csv")
df_202203 <- read_csv("202203-divvy-tripdata.csv")
df_202204 <- read_csv("202204-divvy-tripdata.csv")
df_202205 <- read_csv("202205-divvy-tripdata.csv")
df_202206 <- read_csv("202206-divvy-tripdata.csv")
df_202207 <- read_csv("202207-divvy-tripdata.csv")
df_202208 <- read_csv("202208-divvy-tripdata.csv")
df_202209 <- read_csv("202209-divvy-publictripdata.csv")
```

Using the head function I checked that the headers of each of the dataframes is the same, meaning that appending all of them is possible and will make the exploratory data analysis a lot easier since it will only be done once instead of twelve times.
```{r}
#appends all the dataframes above into one single dataframe
df <- rbind(df_202110, df_202111,df_202112, df_202201, df_202202, df_202203, df_202204, df_202205, df_202206, df_202207, df_202208, df_202209)
```

## Inspecting and cleaning data

The top rows, including headers:
```{r}
head(df)
```
The headers are made of 13 columns, which contain information such as: 
* unique rider IDs, 
* Bike type, 
* start & end times/station name/ID/coordinates, and
* type of rider.

The type of rider will mainly be compared to the other variables since the goal is to understand how members and casual riders use Cyclistic bikes differently. 

We see some missing observations for some of the columns in the head. Using other functions should give more insight into how often they occur and for what columns. But before that we can go ahead and rename the headers in the dataframe to make them more descriptive and cleaner:
```{r}
df_1 <- df %>% 
  rename(Rider_ID = ride_id) %>%
  rename(Bike_Type = rideable_type) %>%
  rename(Rider_Type = member_casual)
```


```{r}
skim_without_charts(df_1)
```
Using the function skim_without_charts from the skimr package shows more information:
* There are over 5 million rows
* Occurances of values in the different rows can be obtained (n_unique). For instance, rideable_type shows that there are three unique types of bikes.
* There are about 5,000 missing values for end_lat and end_lng and about 1,000,000 missing values for the start/end station name/id. In percentages there is a very high percentage of missing data on for start/end station name & ID (about 15%) but also some missing data for the end coordinates (~0.1%). 

Due to the missing values for the geographical data, I will have to make changes to the data. Since locations of the stations should correspond to their respective coordinates, I'll see if there is a possibility to to fill the missing stations with the corresponding station name obtained from matching station-coordinate pairs. 

```{r}
# length(unique(df$column)) gives unique values for the column
length(unique(df_1$start_station_name))
length(unique(df_1$start_station_id))
length(unique(df_1$start_lat))
length(unique(df_1$start_lng))
length(unique(df_1$end_station_name))
length(unique(df_1$end_station_id))
length(unique(df_1$end_lat))
length(unique(df_1$end_lng))
```
There's a large difference in the unique values of station names and IDs to the coordinates, meaning that it will be more complicated to couple the two types of data together than to just equate one value to another. Since I haven't yet learned to use geographical data and I don't see a way to connect the geographical coordinates I will put aside this problem for now by removing the missing rows of coordinates and station names and IDs, while keeping in mind that the data is incomplete.

When looking at the unique values I see that there are more unique station names than unique station IDs. This might be something to keep in mind, but for simplicity I will only use the station names.

```{r}
df_nocoordinates <- df_1 %>%
  select(Rider_ID, Bike_Type, start_station_name, end_station_name, started_at, ended_at, Rider_Type) %>%
  na.omit(df_nocoordinates)
```

## Adding calculated columns
Now add another column that shows the total time the riders used the bikes for, buy taking the difference between 'started_at' and 'ended_at' and calling the new column header 'Total_Time'. Since the result will be expressed in seconds I will change it to minutes by dividing by 60. I will also change the data type to numeric so that it can be used in graphs. I'll change the title of the column to reflect these changes. And I will remove the columns that won't be used at the same time:

```{r}
df_nocoordinates2 <- df_nocoordinates %>%
  mutate(Total_Time_Minutes = (ended_at - started_at)/60) %>%
  select(Rider_ID, Bike_Type, start_station_name, end_station_name, Total_Time_Minutes, Rider_Type)
```

```{r}
df_nocoordinates2$Total_Time_Minutes <- as.numeric(df_nocoordinates2$Total_Time_Minutes)
```

```{r}
head(df_nocoordinates2)
```

Looking at the station names I notice that sometimes riders start and end at the same station. By using this information and comparing it with other data we might see a pattern in rider behavior. I'll add one more column to show when the start and end station is the same or not using Boolean logic:

```{r}
df_nocoordinates3 <- df_nocoordinates2 %>%
  mutate(Round_Trip = start_station_name == end_station_name)
```

```{r}
head(df_nocoordinates3)
```


# Data 

## Bike Type


```{r}
  ggplot(data=df_nocoordinates3) + geom_bar(mapping = aes(x=Rider_Type))
```
There are more annual than casual riders per ride.



```{r}
ggplot(data = df_nocoordinates3) +
  geom_bar(mapping = aes(x = Rider_Type)) +
  facet_wrap(~Bike_Type) +
   labs(title="Bike types used by type of rider", x = "", y = "Count of rides")
```
We see that that there are no docked bikes used at all by annual members.  We also see that the percentage of classic bikes used by members is much greater than it is for casual riders. Casual riders seem to prefer electric bikes over other types of bikes.

Some questions that pop up from this first graph are:
* Why don't any annual members use docked bikes? Are docked bikes available in the areas where they would be used?
* Why do casual riders prefer electric bikes?

## Total Ride Time

```{r scatterplot}
ggplot(data = df_nocoordinates3) +
  geom_point(mapping = aes(x = Total_Time_Minutes, y = Bike_Type)) +
  facet_wrap(~Rider_Type) + 
  labs(title="Total time per ride", x = "Minutes", y = "")
```
It appears that the ride time doesn't seem to differ much between the members and casual riders for electric and classic bikes, although the range of times for the classic bikes is slightly greater than for electric bikes.

However, the docked bikes (which are only used by casual riders) have a a range of usage times that are much greater than either electric or classic bikes.

```{r}
ggplot(data = df_nocoordinates3, aes(x = Total_Time_Minutes, y = Round_Trip)) +
  facet_wrap(~Bike_Type) +
  geom_point() +
  labs(title="Total time per ride", x = "Minutes", y = "")
```

To get a better sense of the spread of the times are we can put the graph above into a boxplot. The 

```{r}
# https://ggplot2.tidyverse.org/reference/geom_density.html
ggplot(data = df_nocoordinates3, aes(Total_Time_Minutes, after_stat(count), fill = Round_Trip)) +
  geom_density(position = "stack")
```


## Geographical Location (Starting location)




##Additionally I would also like to plot Member Type vs. Total rides this year, by counting the 

### Testing assumptions on the data
In the provided project description, the following figures were stated:
* Cyclistic have over 5800 bicycles and 600 docking stations
* The majority of bikers opt for traditional bikes; about 8 percent use assistive options.
* About 30% use bikes for work.

## 
I want to remane the column headers and create calculated columns:
* Total ride time using started_at and ended_at
* Unique riders for each type. E.g. what 
* Are there any riders that are using the service as non-members a lot? What are these users doing?

# Evaluation and concluding remarks

The lack of data on station names (of about 15%) and geographical coordinates (less than 1%) can mean that the results are skewed, especially if the lack of data for these observations are not spread evenly. For example, if there are specific stations that don't show the data, then there could be a significant difference from the reported and actual number of rides in that area.

Recommendations:
* Look into why docked bike users aren't members.